Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-23 paper.tex
First para for GPU section

To demonstrate the ease and efficiency with this data stored in Zarr 
format can be processed on a GPU, we developed two proof-of-concept
applications. In the first we re-implement the af-dist calculation
used throughout as an example of a non-trivial whole-matrix operation
on GPU. In the second, we demonstrate some emerging technology 
in which Zarr data can be decompressed directly in GPU memory. 
In both cases we demonstrate on chromosome 22 of 
1000 Genomes WGS data~\citep{byrska2022high}.

[Details of af-dist benchmark]

[Details of decompression benchmark]

>>>>

To demonstrate the ease and efficiency with this data stored in Zarr 
format can be processed on a GPU, we developed two proof-of-concept
applications. In the first we re-implement the af-dist calculation
used throughout as an example of a non-trivial whole-matrix operation
on GPU. In the second, we demonstrate some emerging technology 
in which Zarr data can be decompressed directly in GPU memory. 
In both cases we demonstrate on chromosome 22 of 
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 26,504 variants and 2504 samples [and more details about the zarr].
Benchmarks in this section were performed on a
machine with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

We first illustrate the ease with which computations can be 
be peformed on the GPU, with little or no preprocessing of the 
data. To do this we reimplemented the af-dist calculation
described above using CuPy~\citep{cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are 
performed on the GPU, mostly with minimal changes required 
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the core operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which 
is also very similar to the Numba compiled function used in other
benchmarks. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array, decompress each to a local
NumPy array and then copying the data to GPU memory by creating 
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 0.045s, with an 
overall data transfer time of 0.19s (we discount decompression
time here; see next paragraph). For comparison, 
our standard Numba compiled classification function required 
a total of 1.44s running on a single CPU core.
Another approach in which we load the entire 
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires 
460MiB) and use
more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions, reduces transfer 
time to 0.136s (and processing time to $2.43\times10^{-4}$s.

Chunk decompression comprises a significant fraction of processing 
time [FIXME crossrefs?]
